Weighted Model Counting Without Parameter Variables

### 2021-03-07 doc/paper3/paper.tex
Introduction and conclusion

\begin{abstract}
  The abstract should briefly summarize the contents of the paper in
  150--250 words.

@@

\section{Introduction}

% TODO for later:
% 1) read through v2 of the paper and check if I forgot to mention something
% here
% 2) run this through grammarly

% NOTE: What is the problem? Why is it interesting and important?
WMC has emerged as a powerful computational framework for problems in many
domains, e.g., probabilistic graphical models such as Bayesian networks and
Markov networks
\cite{DBLP:conf/ecai/BartKLM16,DBLP:conf/ijcai/ChaviraD05,DBLP:conf/sat/ChaviraD06,DBLP:conf/kr/Darwiche02,DBLP:conf/aaai/SangBK05},
neuro-symbolic artificial intelligence \cite{DBLP:conf/icml/XuZFLB18},
probabilistic programs \cite{DBLP:journals/pacmpl/HoltzenBM20}, and
probabilistic logic programs \cite{DBLP:journals/tplp/FierensBRSGTJR15}. It has
been extended to support continuous variables \cite{DBLP:conf/ijcai/BellePB15},
infinite domains \cite{DBLP:conf/aaai/Belle17}, first-order logic
\cite{DBLP:journals/cacm/GogateD16,DBLP:conf/ijcai/BroeckTMDR11}, and arbitrary
semirings \cite{DBLP:journals/ijar/BelleR20,DBLP:journals/japll/KimmigBR17}.

% NOTE: What's wrong with previous proposed solutions? How does mine differ?
% TODO: Our work is similar to \cite{DBLP:conf/uai/GogateD10} who suggest using
% a collection of soft (i.e., with weights) and hard clauses
% TODO: how different are the semantics?
% But:
% 1) we can handle arbitrary formulas and not just clauses
% 2) we assign two weights, not one
% 3) the semantics of these weights are completely different, similar to the
% ones from Markov networks. It is thus difficult to reuse the same formulation
% for a different WMC algorithm.
% 4) Since we present a WMC->PBP algorithm, we don't need to invent new
% encodings for each problem and we can benefit from all the good ideas that
% went into designing WMC encodings

Many WMC inference algorithms such as
\textsf{Ace}\footnote{\url{http://reasoning.cs.ucla.edu/ace/}}, \textsf{c2d}
\cite{DBLP:conf/ecai/Darwiche04}, and \textsf{miniC2D}
\cite{DBLP:conf/ijcai/OztokD15} work by compilation to tractable representations
such as arithmetic circuits, deterministic, decomposable negation normal form
\cite{DBLP:journals/jancl/Darwiche01}, and sentential decision diagrams (SDDs)
\cite{DBLP:conf/ijcai/Darwiche11}. Some attempts have previously been made to
skip this intermediate stage of representing a Bayesian network as a WMC
instance and compile directly to a more convenient representation.
Specifically, direct compilation from Bayesian networks to SDDs
\cite{DBLP:conf/ecsqaru/ChoiKD13} and from structured Bayesian networks (i.e., a
generalisation of Bayesian networks) to probabilistic SDDs
\cite{shen2020modeling} have been considered. To the best of the authors'
knowledge, neither compilation approach has a publicly available implementation.

% NOTE: What are the key components of my approach and results? Also include any
% specific limitations. What about your project is new or different from
% existing solutions?
Our work is similar in that it introduces a new representation for computational
problems that is based on pseudo-Boolean functions---pseudo-Boolean projection
(PBP). We formally show that every WMC problem instance has a corresponding PBP
instance. We argue that this new format is more appropriate for WMC solvers
whose execution is build on pseudo-Boolean function manipulation, i.e.,
\textsf{ADDMC} \cite{DBLP:conf/aaai/DudekPV20} and \textsf{DPMC}
\cite{DBLP:conf/cp/DudekPV20}. Finally, we show how most methods of encoding
probabilistic inference on Bayesian networks to WMC can be adapted to the PBP
format. This transformation significantly improves inference speed,
outperforming all other considered algorithm-encoding combinations.

%allowing two encodings that were previously incompatible with
%\textsf{DPMC} to be compatible.

% NOTE: List of refutable contributions (with forward references to the rest of
% the paper) What is the outcome of the project?
\paragraph{Notes.}
\begin{itemize}
\item we also show correctness
\item and introduce a new notation
\item TODO: briefly mention what a pseudo-Boolean function is
\end{itemize}

@@

@@

In this case, we have to add some assumptions about the cardinality of models.
\Cref{cond:22} states that all models of the indicator-only part of the formula
have the same cardinality. Bayesian network encodings such as \texttt{cd05} and
\texttt{cd06} satisfy this condition by assigning an indicator variable to each
possible variable-value pair and requiring each random variable to be paired
with exactly one value. \Cref{cond:23} then says that the smallest number of
parameter variables needed to turn an indicator-only model into a full model is
the same for all indicator-only models. As some ideas duplicate between the
proofs of \cref{thm:correctness,thm:mccorrectness}, the following proof is
slightly less explicit and assumes that $\omega = 1$.

% We assume that $\omega = 1$ (i.e., $\{ p \} \not\in \phi$ for all parameter
% variables $p \in X_P$) as the proof can be easily extended to handle $\omega
% \neq 1$ similarly to the proof of \cref{thm:correctness}.

% TODO: check if each condition is actually used

@@

Overall, transforming WMC instances to the PBP format allows us to significantly
simplify each instance. This transformation is particularly effective on
\texttt{bklm16}, allowing it to surpass \texttt{cd06} and become the new state
of the art. While there is a similarly significant reduction in the number of
variables for \texttt{d02}, the performance of $\textsf{DPMC}+\texttt{d02}$ is
virtually unaffected. Finally, while our transformation makes it possible to use
\texttt{cd05} and \texttt{cd06} with \textsf{DPMC}, the two combinations remain
inefficient.

\section{Conclusions}

\paragraph{Notes.}
\begin{itemize}
\item Benefits of my approach:
  \begin{itemize}
  \item smaller primal graph, so easier to perform tree decomposition
  \item Variable order is less likely to be obstructed by all the unnecessary
    'parameter' variables.
  \item There are others, but they're not that important.
  \end{itemize}
\item Future work: do other WMC encodings also have this variable partition
  property?
\item Do I need to formally consider extending a pseudo-Boolean function to a
  bigger domain?
\item Benefits of having this proof in the paper:
  \begin{itemize}
  \item It puts all encodings on a common ground.
  \item It illustrates the convenience of our notation for reasoning about
    (certain types of) pseudo-Boolean functions.
  \item It's too big and too important to be left for the appendix.
  \end{itemize}
\end{itemize}

>>>>

\begin{abstract} % TODO: write
  The abstract should briefly summarize the contents of the paper in
  150--250 words.

@@

\section{Introduction}
% TODO for later:
% 1) read through v2 of the paper and check if I forgot to mention something
% here
% 2) run this through grammarly

% NOTE: What is the problem? Why is it interesting and important?
Weighted model counting (WMC), i.e., a generalisation of propositional model
counting that assigns weights to literals and computes the total weight of all
models of a propositional formula \cite{DBLP:journals/ai/ChaviraD08}, has
emerged as a powerful computational framework for problems in many domains,
e.g., probabilistic graphical models such as Bayesian networks and Markov
networks
\cite{DBLP:conf/ecai/BartKLM16,DBLP:conf/ijcai/ChaviraD05,DBLP:conf/sat/ChaviraD06,DBLP:conf/kr/Darwiche02,DBLP:conf/aaai/SangBK05},
neuro-symbolic artificial intelligence \cite{DBLP:conf/icml/XuZFLB18},
probabilistic programs \cite{DBLP:journals/pacmpl/HoltzenBM20}, and
probabilistic logic programs \cite{DBLP:journals/tplp/FierensBRSGTJR15}. It has
been extended to support continuous variables \cite{DBLP:conf/ijcai/BellePB15},
infinite domains \cite{DBLP:conf/aaai/Belle17}, first-order logic
\cite{DBLP:journals/cacm/GogateD16,DBLP:conf/ijcai/BroeckTMDR11}, and arbitrary
semirings \cite{DBLP:journals/ijar/BelleR20,DBLP:journals/japll/KimmigBR17}.
However, as the definition of WMC puts weights on literals, additional variables
often need to be added for the sole purpose of holding a weight
\cite{DBLP:conf/ecai/BartKLM16,DBLP:conf/ijcai/ChaviraD05,DBLP:conf/sat/ChaviraD06,DBLP:conf/kr/Darwiche02,DBLP:conf/aaai/SangBK05}.
As the parameterised complexity of model counting (and, by extension, WMC)
depends on the number of variables
\cite{DBLP:conf/focs/BacchusDP03,DBLP:journals/jda/SamerS10}, this can make WMC
unnecessarily slow and could be detrimental to WMC algorithms such as
\textsf{ADDMC} \cite{DBLP:conf/aaai/DudekPV20} that depend on variable ordering
heuristics.

% NOTE: What's wrong with previous proposed solutions? How does mine differ?
One approach to this problem considers weighted clauses and probabilistic
semantics based on Markov networks \cite{DBLP:conf/uai/GogateD10}. However, with
a new representation comes the need to invent new encodings and inference
algorithms. Our work is similar in that it introduces a new representation for
computational problems but is able to reuse recent WMC algorithms based on
pseudo-Boolean function manipulation, namely, \textsf{ADDMC}
\cite{DBLP:conf/aaai/DudekPV20} and \textsf{DPMC} \cite{DBLP:conf/cp/DudekPV20}.
Furthermore, we identify necessary conditions for transforming a WMC instance
into the new format. As many WMC inference algorithms such as
\textsf{Ace}\footnote{\url{http://reasoning.cs.ucla.edu/ace/}}, \textsf{c2d}
\cite{DBLP:conf/ecai/Darwiche04}, and \textsf{miniC2D}
\cite{DBLP:conf/ijcai/OztokD15} work by compilation to tractable representations
such as arithmetic circuits, deterministic, decomposable negation normal form
\cite{DBLP:journals/jancl/Darwiche01}, and sentential decision diagrams (SDDs)
\cite{DBLP:conf/ijcai/Darwiche11}, another way to avoid parameter variables
could be via direct compilation to a more convenient representation. However,
while direct compilation of Bayesian networks to SDDs has been investigated
\cite{DBLP:conf/ecsqaru/ChoiKD13}, SDDs are not expressive enough to avoid the
issue.

% NOTE: What are the key components of my approach and results? Also include any
% specific limitations. What about your project is new or different from
% existing solutions? List of refutable contributions (with forward references
% to the rest of the paper) What is the outcome of the project?
In this work, we introduce a way to transform WMC problems to a new format based
on pseudo-Boolean functions---\emph{pseudo-Boolean projection} (PBP). We
formally show that every WMC problem instance has a corresponding PBP instance
and identify conditions under which this transformation can remove parameter
variables. Four out of the five known WMC encodings for Bayesian networks
\cite{DBLP:conf/ecai/BartKLM16,DBLP:conf/ijcai/ChaviraD05,DBLP:conf/sat/ChaviraD06,DBLP:conf/kr/Darwiche02,DBLP:conf/aaai/SangBK05}
can indeed be simplified in this manner. This transformation enables two
encodings that were previously incompatible with most WMC algorithms (due to
using a different definition of WMC
\cite{DBLP:conf/ijcai/ChaviraD05,DBLP:conf/sat/ChaviraD06}) to be run with
\textsf{ADDMC} and \textsf{DPMC} and results in a significant performance boost
for one other encoding, allowing it to surpass the current state of the art.
Finally, our theoretical contributions result in a convenient algebraic way of
reasoning about two-valued pseudo-Boolean functions and position WMC encodings
on a common ground, identifying their key properties and assumptions.

%To the best of the authors' knowledge, none of these compilation/inference
%techniques have a publicly available implementation.

% Specifically, direct compilation
% from Bayesian networks to SDDs \cite{DBLP:conf/ecsqaru/ChoiKD13} and from
% structured Bayesian networks (i.e., a generalisation of Bayesian networks) to
% probabilistic SDDs \cite{shen2020modeling} have been considered.

% * SDD compilation still has parameter variables
% 1) However, no conversion algorithms from either WMC or probabilistic models
% are provided.
% Since we present a WMC to PBP algorithm, we don't need to invent new
% encodings for each problem and we can benefit from all the good ideas that
% went into designing WMC encodings
% 3) Instead of combining techniques from older WMC algorithms like they did, we
% use a much more recent WMC solver that is shown to...
% 3) We benefit from a decade of research in WMC inference and, specifically, 
% 4) We are not restricted to clauses and probabilistic semantics
% 5) We identify conditions and describe an algorithm for transforming any WMC
% instance into PBP

@@

% TODO: briefly mention what a pseudo-Boolean function is
% TODO: mention that $2^x$ denotes the powerset of $X$.

@@

In this case, we have to add some assumptions about the cardinality of models.
\Cref{cond:22} states that all models of the indicator-only part of the formula
have the same cardinality. Bayesian network encodings such as \texttt{cd05} and
\texttt{cd06} satisfy this condition by assigning an indicator variable to each
possible variable-value pair and requiring each random variable to be paired
with exactly one value. \Cref{cond:23} then says that the smallest number of
parameter variables needed to turn an indicator-only model into a full model is
the same for all indicator-only models. As some ideas duplicate between the
proofs of \cref{thm:correctness,thm:mccorrectness}, the following proof is
slightly less explicit and assumes that $\omega = 1$.
% We assume that $\omega = 1$ (i.e., $\{ p \} \not\in \phi$ for all parameter
% variables $p \in X_P$) as the proof can be easily extended to handle $\omega
% \neq 1$ similarly to the proof of \cref{thm:correctness}.

@@

Overall, transforming WMC instances to the PBP format allows us to significantly
simplify each instance. This transformation is particularly effective on
\texttt{bklm16}, allowing it to surpass \texttt{cd06} and become the new state
of the art. While there is a similarly significant reduction in the number of
variables for \texttt{d02}, the performance of $\textsf{DPMC}+\texttt{d02}$ is
virtually unaffected. Finally, while our transformation makes it possible to use
\texttt{cd05} and \texttt{cd06} with \textsf{DPMC}, the two combinations remain
inefficient.
% TODO: mention (somewhere) that we can't compare to Gogate's work and direct
% compilation (take the text from the introduction)

\section{Conclusion} % So what?

In this paper, we showed how the number of variables in a WMC instance can be
significantly reduced by transforming it into a representation based on
two-valued pseudo-Boolean functions. In some cases, this led to significant
improvements in inference speed, allowing $\textsf{DPMC} + \texttt{bklm16++}$ to
overtake $\textsf{Ace}+\texttt{cd06}$ as the new state of the art WMC technique
for Bayesian network inference. Moreover, we identified key properties of
Bayesian network encodings that allow for parameter variable removal. However,
these properties were rather different for each encoding, and so an interesting
question for future work is whether they can be unified into a more abstract
and coherent list of conditions.

Bayesian network inference was chosen as the example application of WMC because
it is the first and the most studied one
\cite{DBLP:conf/ecai/BartKLM16,DBLP:conf/ijcai/ChaviraD05,DBLP:conf/sat/ChaviraD06,DBLP:conf/kr/Darwiche02,DBLP:conf/aaai/SangBK05}.
While the distinction between indicator and parameter variables is often not
explicitly described in other WMC encodings
\cite{DBLP:journals/tplp/FierensBRSGTJR15,DBLP:journals/pacmpl/HoltzenBM20,DBLP:conf/icml/XuZFLB18},
perhaps in some cases variables could still be partitioned in this way, allowing
for not just faster inference with \textsf{DPMC} or \textsf{ADDMC} but also for
well-established WMC encoding and inference techniques (such as in the
\texttt{cd05} and \texttt{cd06} papers
\cite{DBLP:conf/ijcai/ChaviraD05,DBLP:conf/sat/ChaviraD06}) to be transferred to
other application domains.
